---
title: "Heart Attack Visualisation"
output: html_notebook
---
Challenge: What key variables relate to low versus high risk of heart attack. 

## Load in data and packages
```{r message=FALSE}
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(wesanderson)
library(GGally)
library(tableone)
library(corrplot)

df <- read.csv("/Users/rebecca/Desktop/heart.csv")
# Print first 10 lines to see data and columns
head(df, 10)
```

## Data Description:
- Age: Numeric (e.g., 52) 
- Sex: Categorical (0: Female, 1: Male) 
- Chest Pain Type: Categorical (0: Typical Angina, 1: Atypical Angina, 2: Non-anginal Pain, 3: Asymptomatic) 
- Resting Blood Pressure: Numeric (e.g., 125) 
- Serum Cholesterol: Numeric in mg/dL (e.g., 212) 
- Fasting Blood Sugar: Categorical (0: <= 120 mg/dL, 1: > 120 mg/dL) 
- Resting Electrocardiographic Results: Categorical (0: Normal, 1: Abnormality, 2: Hypertrophy) 
- Maximum Heart Rate Achieved: Numeric (e.g., 168) 
- Exercise-Induced Angina: Categorical (0: No, 1: Yes) 
- Oldpeak (ST Depression): Numeric (e.g., 1.0) 
- Slope of Peak Exercise ST Segment: Categorical (0: Upsloping, 1: Flat, 2: Downsloping) 
- Number of Major Vessels Colored by Fluoroscopy: Numeric (0 to 3) 
- Thalassemia: Categorical (0: Normal, 1: Fixed Defect, 2: Reversible Defect)

Key outcome:
- "Target": the percentage of heart attack risk in patients.

## Data Cleaning:
```{r}
# Edit data types e.g. sex as categorical
cols <- c("sex", "cp", "fbs", "restecg", "exang", "slope", "ca", "thal")
df[,cols] <- lapply(df[,cols], factor) 

# Check for missing data:
miss <- sum(is.na(df))
sprintf("Number of missing datapoints: %i", miss)

# Check for duplicate rows & remove
dup <- sum(duplicated.data.frame(df))
sprintf("Number of duplicated rows: %i", dup)
if(dup > 0){
  df <- distinct(df)
}
```

# Summary Statistics
```{r}
CreateTableOne(data=df, factorVars = cols)

# Plot Target as histogram
ggplot(data=df, aes(x=target))+
  geom_histogram(color="darkblue", fill="lightblue")

# Clear grouping of 'high' and 'low' risk groups for heart attack 
df$target_factor <- factor(ifelse(df$target <0.5, 'low', 'high'), levels=c('low','high')) 
```

## Plot continuous and categorical variables and how they relate to target
```{r message=FALSE}
# How does each continuous variable relate to 'target_factor'
contin <- c("age", "trestbps", "chol", "thalach", "oldpeak")
df[,c(contin, "target_factor")] %>%
  gather(-target_factor, key = "var", value = "value") %>%
  ggplot(aes(y = value, x = target_factor, color= target_factor)) +
    geom_boxplot() +
    facet_wrap(~ var, scales = "free")

# How does each categorical variable relate to 'target'
df[,c(cols, "target")] %>%
  gather(-target, key = "var", value = "value") %>%
  ggplot(aes(x = value, y = target, group= value, fill=value)) +
    geom_boxplot() +
    facet_wrap(~ var, scales = "free")
```

## Correlation plot between all variables and target outcome
```{r}
# mat : is a matrix of data
# ... : further arguments to pass to the native R cor.test function
cor.mtest <- function(mat, ...) {
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat<- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}
# matrix of the p-value of the correlation
M <- cor(df[,c(contin, "target")])
p.mat <- cor.mtest(df[,c(contin, "target")])

# Leave blank on no significant coefficient
corrplot(M, type="upper", order="hclust", 
         p.mat = p.mat, sig.level = 0.01, insig = "blank")
```

# Which of these are significantly related to low/high risk groups?
## Statistical Tests
```{r}
# For continuous data, perform a t-test between low and high risk groups
stat1 = data.frame(matrix(ncol = 4, nrow = length(contin)))
colnames(stat1) <- c("Var", "Test", "Test_Stat", "p")
stat1$Var <- contin
stat1$Test <- "t-test"

for (i in 1:length(contin)){
  var <- contin[i]
  test <- t.test(df[,var] ~ df$target_factor)
  stat1$Test_Stat[i] <- test$statistic
  stat1$p[i] <- test$p.value
}

# For categorical data, use a chi-square test 
stat2 = data.frame(matrix(ncol = 4, nrow = length(cols)))
colnames(stat2) <- c("Var", "Test", "Test_Stat", "p")
stat2$Var <- cols
stat2$Test <- "chi-sq"

for (i in 1:length(cols)){
  var <- cols[i]
  test <- chisq.test(df[,var], df$target_factor, correct=FALSE)
  stat2$Test_Stat[i] <- test$statistic
  stat2$p[i] <- test$p.value
}


# Combined these two tables and add fdr-correction for number of tests being performed, and add significance indicator
stat3 <- rbind(stat1, stat2)
stat3$p_fdr <- p.adjust(stat3$p, method="fdr")
stat3$sig <- cut(stat3$p_fdr, breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1), include.lowest = T, labels = c('***', '**', '*', '.', ''))
stat3
```

# Key Takeaways:
* Clear separation between low and high risk patients
* This may be more clinically useful as an indicator, versus absolute risk percentage
* All variables, excpet for fasting blood sugar, show significant relationships to high versus low risk
* These are viable candidates for a predictive model to determine future risk status of patients 

